### 2017-02-07 manuscript.rmd
added pdf

---
bibliography: manuscript.bib
csl: nature.csl
output:
  github_document: default
  pdf_document:
    dev: png
---

@@

A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of trait comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew]. A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share many traits that evolve once in a common ancestor. This violates the fundamental assumption of observation independence in standard statistical methods. Phylogenetic comparative methods address this dependence. They have largely been applied to morphological and ecological traits, but are just as relevant to functional genomics. Even so, most comparative functional genomics studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons (Fig. 1A). This leaves comparative functional genomics studies susceptible to statistical problems and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.

Phylogenetic comparative methods account for evolutionary history and explicitly model trait change along the branches of evolutionary trees. The value of these methods relative to to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq]. The application of these methods is illustrated by a study of leaf functional attributes for about 100 plant species [@Ackerly:1999iu]. If phylogenetic relationships are not considered, analyses indicate a negative correlation of leaf lifespan with leaf size. With the application of phylogenetic comparative methods, however, this correlation disappears. This is because most of the observed variance is due to differences between conifers and flowering plants. Phylogenetic comparative methods capture this shift as change along a single branch deep in the tree, revealing that there is not a tendency of correlated change between these traits across the phylogeny. One reason that comparative functional genomics has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomics data. Here we examine this issue be re-evaluating the pairwise comparisons used in two recent studies that compared gene expression across species.

The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend to have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm, @Koonin:2005fp]. Using pairwise comparisons (Fig. 1A), they found lower expression correlation between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.

The second comparative functional genomics study we evaluate here is Levin *et al.* [@Levin:2016bp]. This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum. Using pairwise comparisons (Figure 1A), they found there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases (Fig. XX). They suggest that this supports an "inverse hourglass" model for the evolution of gene expression, in contrast with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw], and provides biological justification for the concept of phyla. We previously described concerns with the interpretations of this result [@Hejnol:2016hw]. Here we address the analyses themselves by examining the structure of the pairwise comparisons.

![Figure Overview](Figure_overview.png?raw=true)

> Figure 1 | Pairwise and phylogenetic comparative approaches. (a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes (such as orthologs and paralogs). This leads to many more comparisons than unique observations, making each comparison dependent on others. (b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches. Phylogenetic approaches are rarely used for functional genomics studies.

@@

The KMRR analyses are based on pairwise comparisons (Figure 1A) between Tau within each gene family. They found the correlation coefficient of Tau for orthologs to be significantly greater than the correlation coefficient of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs. From this they concluded that their analyses support the ortholog conjecture. They also concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs" [@KryuchkovaMostacci:2016iw].

@@

```{r expectation_plots, echo=FALSE, cache=TRUE, message=F, warning=F}

@@

> Figure XXExpectations | (a) Under the null hypothesis that there is no difference in the rate of evolution after duplication or speciation events, the distribution of PICs would be the same across both types of nodes. (b) Under the ortholog conjecture, contrasts across duplication nodes would tend to be larger than contrasts across speciation nodes.

For each internal node in each gene tree, we measured the phylogenetic independent contrast (PIC) [@Felsenstein:1985ua] of the expression statistic Tau. This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Figure 1b). These contrasts were then annotated by whether each is made across a speciation or duplication event. The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree. Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree. We present eight of the actual gene trees with Tau values and independent contrasts in Figure XX_EXAMPLE_TREES as an example of how these methods are applied.

@@

```{r show_contrast_result, echo=FALSE, cache=TRUE, message=F, warning=F}

@@

> Figure XXKMRR | Density plot of the magnitude of phylogenetic independent contrast values following duplication and speciation events. These contrasts are not larger following duplication events, as predicted by the ortholog conjecture.

We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] datset onto `r length(gene_trees)` gene trees parsed from ENSMBL Compara [@Herrero:2016jj]. These are the same pre-computed trees that the orthology/ paralogy annotations KMRR used are based on. `r length(gene_trees_pruned)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values and had at least one speciation event. Of these, `r length(gene_trees_calibrated)` were successfully time calibrated. These calibrated trees were used to calculate phylogenetic independent contrasts for  `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes.

It is essential to have a null hypothesis that makes a distinct prediction from the prediction of the hypothesis under consideration. A suitable null hypothesis in this case  is that there is no difference in the evolution of expression following speciation or duplication events [@Studer:2009bm]. Under this hypothesis, we would predict that contrasts across speciation nodes and duplication nodes are drawn from the same distribution (Figure XXEXPECTATIONS a). Under the alternative hypothesis specified by the ortholog conjecture, that there is a higher rate of change following duplication events than speciation events, we would expect to see the distribution of duplication contrasts shifted to higher values relative to the speciation contrasts (Figure XXEXPECTATIONS b).

We did not find increased evolutionary change in expression following duplication events (Fig. XXKMRR). The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result`). Our phylogenetic comparative analysis, unlike the previously published pairwise comparative analysis [@KryuchkovaMostacci:2016iw], therefore find no evidence for the ortholog conjecture in this system.

@@

We next examined the possibility that ascertainment biases were differentially impacting the inference of expression evolution following duplication and speciation events. Such a bias might obscure support for the ortholog conjecture. We focused on two possible sources of bias, node depth and branch length. We found no evidence that either affected our results (Supplementary Materials). We also examined the sensitivity of the results to the calibration times applied to speciation events on the gene trees. This is important because it is expected that genes from separate species have a common ancestor older than the time at which the species diverged from each other [@Takahata:1989uh; @Degnan:2009hr]. There is also uncertainty associated with the timing of these speciation events. We added random noise to the calibration times in replicate analyses, and all still failed to reject the null hypothesis (Supplementary Materials).

@@

```{r pairwise, echo=FALSE, cache=TRUE, message=F, warning=F}
	get_pairwise_summary = function ( nhx ) {

@@

	pairwise_summary %>% 
		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
		filter( ! is.na( D ) ) %>%
		group_by( Event, label ) %>% 
		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
		rename( clade = label ) %>% 
		left_join( calibration_times, clade=clade ) %>%
		ggplot() + 
			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
			theme_classic() +
			xlab( "Node Age (million years)" ) +
			ylab( "Correlation Coefficient" )

@@

> Figure XX_KMRR_Pairwise | This figures demonstrates that the pairwise analyses presented in Figure 2a of KMRR can be reproduced with the subset of data we consider here and our own independent code base. Each circle indicates the Tau Pearson correlation coefficient for a set of pairwise comparisons annotated with a specific node name of a given age and event type, *i.e.*, whether the divergence at that nodes was due to speciation (giving rise to orthologs) or duplication (giving rise to paralogs). Following KMRR Figure 2a, this figure shows only the comparisons that include a human sequence.

@@

Why is it that pairwise methods and phylogenetic methods lead to opposite conclusions? The reason is that they describe different things. Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes [@OMeara:2012gq]. There need not be a different process of evolution following speciation and duplication for paralogs to be more different than orthologs. This could be due to the structure of the gene phylogenies. If paralogs tend to be more distantly related to each other than orthologs, then there would be more time for differences to accumulate even if the rate of change is the same between the two. This is, in fact, the case for these data. While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` million years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` million years. This is because the oldest speciation event is by definition the most recent common ancestor of the species included in the study, but many gene families underwent duplication before this time.

To test the hypothesis that distantly paralogs that arose from duplications that precede the oldest speciation event (Supplementary Figure XXAscertainment a) impact the lower correlation of Tau between paralogs than between orthologs, we removed them. When we consider only the paralogs that arise from duplication events the same age or younger than the oldest speciation event (Supplementary Figure XXAscertainment b), the paralog correlation coefficient increases from `r round( paralog_r, 3)` to  `r round( paralog_subset_r, 3)`. This is much closer to the ortholog correlation of `r round( ortholog_r, 3) `.

@@

```{r simulations_plot, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}

@@

	g_pair_null = 
		pairwise_sim_null_summary %>% 
		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
		filter( ! is.na( D ) ) %>%
		group_by( Event, label ) %>% 
		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
		rename( clade = label ) %>% 
		left_join( calibration_times, clade=clade ) %>%
		ggplot() + 
			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
			theme_classic() +
			xlab( "Node Age (million years)" ) +
			ylab( "Correlation Coefficient" ) +
			ylim( 0, 1 ) +
			ggtitle( "(a) Null simulation, pairwise" ) +
			theme(legend.position="none")

@@

	g_pair_oc = 
		pairwise_sim_oc_summary %>% 
		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
		filter( ! is.na( D ) ) %>%
		group_by( Event, label ) %>% 
		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
		rename( clade = label ) %>% 
		left_join( calibration_times, clade=clade ) %>%
		ggplot() + 
			geom_point( aes( x=age, y=rho, col=Event, size=n ) ) +
			theme_classic() +
			xlab( "Node Age (million years)" ) +
			ylab( "Correlation Coefficient" ) +
			ylim( 0, 1 ) +
			ggtitle( "(c) OC simulation, pairwise" ) +
			theme(legend.position="none")

@@

> Figure XXSim

The KMRR study did investigate the impact of node age on correlation, but in a different way. In their Figure 2, they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age. They found that across the investigated range of node ages, ortholog pairs have higher tau correlation than paralogs. We confirmed that we can replicate this result (Figure XX_KMRR_Pairwise). There are several difficulty with interpreting this plot, though. First, it does not just reflect the evolutionary processes that generated the data, it is also impacted by the phylogenies along which these processes acted. The expected covariance of traits that evolve under neutral processes is in fact defined by the phylogeny [@Grafen:1989um]. Second, the correlation for each group is based on multiple non-independent pairwise comparisons.

To better understand this plot (Figure XX_KMRR_Pairwise), we performed a couple simulations of Tau on the calibrated gene trees and then regenerated the figures. First, we simulated the evolution of tau under the null model that it evolves at the same rate following duplication and speciation events. Under the null model, the plot of mean correlation to node age is very similar as for the observed data (Figure XXSim a). As in the original study, there is higher correlation coefficient across orthologs (`r ortholog_r_sim_null`) than paralogs (`r paralog_r_sim_null`) when not considering node age. Phylogenetic analysis of the data simulated under the null hypothesis (Figure XXSim b) do not reject the null hypothesis (Wilcoxon p=`r wilcox_oc( nodes_sim_null_contrast )`), as expected.

We next simulated the evolution of Tau under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation. The pairwise results of this heterogeneous model are nearly indistinguishable from the null model (Figure XXSim c), and also have a higher correlation coefficient for orthologs (`r ortholog_r_sim_oc`) than paralogs (`r paralog_r_sim_oc`). The phylogenetic analysis of the ortholog conjecture simulation (Figure XXSim c) does reject the null hypothesis (Wilcoxon p=`r wilcox_oc( nodes_sim_oc_contrast )`).

@@

Their Figure 3 differs from their analyses in that it presents independent changes in expression between triplets of genes. Each triplet has two paralogs that arose from a duplication event and one unduplicated homolog. They find that there is a tendency for the paralog with the lowest expression to have the highest Tau. From this they conclude that as paralogs diverge in expression, there is a tendency for the one that evolves to have lower expression also evolve to become more tissue specific. There is, however, a negative relationship between Tau and maximum expression across all genes (Supplementary Figure XX_TAU_V_EXP). A simpler explanation for this plot is that it reflects the global tendency for Tau to decrease with maximum expression. This global pattern could be due to both biological factors and the technical details of how Tau is calculated.

@@

![Figure Levin](levin_etal/Figure_levin.png?raw=true)

> Figure XXLevin | Distributions of pairwise similarity scores for each phase of development. Pairwise scores for the ctenophore are red. Wilcoxon test p-values for the significance of the differences between early-mid distributions and late-mid distributions are on the right. Model of variance, which is inversely related to similarity, is on the left. (a) The distributions as published. Low similarity (*i.e.*, high variance) in the mid phase of development was interpreted as support for an inverse hourglass model for the evolution of gene expression. The five least-similar mid phase scores were all from the ctenophore. Published KS p-values, based on duplicated data, are in parentheses. The inset ctenophore image is by S. Haddock from phylopic.org. (b) The distributions after the exclusion of the ctenophore. The early and mid phase distributions are not statistically distinct. This suggest a wine bottle model, with similar evolutionary variance at the early and mid phase and less at the late phase.

@@

We examined the matrix of pairwise comparisons used as the base for the KS tests and Figure 4d in *Levin et al.* [@Levin:2016bp], and thus as evidence to support the "inverse hourglass model". We found several problems resulting from the use of multiple pairwise comparisons. The first problems are specific to this particular implementation of pairwise comparisons. We found that every data point was included twice because both reciprocal pairwise comparisons (which have the same values) were retained. For example, there is both a nematode to arthropod comparison and an arthropod to nematode comparison. As a consequence, there are 90 entries for the 45 pairwise comparisons, and by doubling the data the significance of the result appears stronger than it actually is. After removing the duplicate values, the p values are far less significant, 0.002 for the early-mid comparison and on the order of 10<sup>-6</sup> for early-late. In addition, the test they used (KS test) is not appropriate for the hypothesis they seek to evaluate. The KS test does not just evaluate whether one distribution is greater than the other, it also tests whether the shape of the distributions are the same. In addition, the samples in this dataset are matched (*i.e.*, for each pairwise comparison there is a early, mid, and late expression value), which the KS test does not take into account. The Wilcoxon test is instead appropriate in this case. When applied to the de-duplicated data, the significance of this test is 0.02 for the early-mid comparison and on the order of 10<sup>-7</sup> for early-late.

>>>>

---
bibliography: manuscript.bib
csl: nature.csl
output:
  pdf_document:
    dev: png
    fig_caption: yes
    fig_width: 6
    fig_height: 4
  github_document:
    fig_caption: yes
header-includes:
  - \setlength{\parindent}{4em}
---

@@

A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of trait comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew]. A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share many traits that evolved once in a common ancestor. This violates the fundamental assumption of observation independence in standard statistical methods. Phylogenetic comparative methods address this dependence. They have largely been applied to morphological and ecological traits, but are just as relevant to functional genomics. Even so, most comparative functional genomics studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons (Figure 1a). This leaves comparative functional genomics studies susceptible to statistical problems and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.

Phylogenetic comparative methods account for evolutionary history and explicitly model trait change along the branches of evolutionary trees (*e.g.*, Figure 1b). The value of these methods relative to to pairwise comparisons has been repeatedly shown in analyses of other types of character data [@Ricklefs:1996wb; @Chamberlain:2012vx; @OMeara:2012gq]. The application of these methods is illustrated by a study of leaf functional attributes for about 100 plant species [@Ackerly:1999iu]. If phylogenetic relationships are not considered, analyses indicate a negative correlation of leaf lifespan with leaf size. With the application of phylogenetic comparative methods, however, this correlation disappears. This is because most of the observed variance is due to differences between conifers and flowering plants. Phylogenetic comparative methods capture this shift as change along a single branch deep in the tree, revealing that there is not a tendency of correlated change between these traits across the phylogeny. One reason that comparative functional genomics has not yet embraced phylogenetic approaches is that it has not yet been concretely demonstrated that pairwise and phylogenetic comparative methods can lead to different results when considering functional genomics data. Here we examine this issue be re-evaluating the pairwise comparisons used in two recent studies that compared gene expression across species.

The first study, Kryuchkova-Mostacci and Robinson-Rechavi (KMRR) 2016 [@KryuchkovaMostacci:2016iw], analyzed multiple vertebrate expression datasets to test the ortholog conjecture - the hypothesis that orthologs tend to have more conserved attributes (specificity of expression across organs in this case) than do paralogs [@Nehrt:2011bm, @Koonin:2005fp]. Using pairwise comparisons (Figure 1a), they found lower expression correlation between paralogs than between orthologs and interpreted this as strong support for the ortholog conjecture.

The second comparative functional genomics study we evaluate here is Levin *et al.* [@Levin:2016bp]. This study analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum. Using pairwise comparisons, they found there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases (Figure 1a). They suggest that this supports an "inverse hourglass" model for the evolution of gene expression, in contrast with the "hourglass" model previously proposed for closely related species [@Kalinka:2010dw], and provides biological justification for the concept of phyla. We previously described concerns with the interpretations of this result [@Hejnol:2016hw]. Here we address the analyses themselves by examining the structure of the pairwise comparisons.

![Pairwise and phylogenetic comparative approaches. (a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes (such as orthologs and paralogs). This leads to many more comparisons than unique observations, making each comparison dependent on others. (b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches. Phylogenetic approaches are rarely used for functional genomics studies.](Figure_overview.png?raw=true)

@@

The KMRR analyses are based on pairwise comparisons (Figure 1a) between Tau within each gene family. They found the correlation coefficient of Tau for orthologs to be significantly greater than the correlation coefficient of Tau for paralogs, *i.e.* that orthologs tend to be more similar to each other than paralogs. From this they concluded that their analyses support the ortholog conjecture. They also concluded that this pattern provides support for a particular evolutionary process, that "tissue-specificity evolves very slowly in the absence of duplication, while immediately after duplication the new gene copy differs" [@KryuchkovaMostacci:2016iw].

@@

```{r expectation_plots, echo=FALSE, cache=TRUE, message=F, warning=F, fig.cap="(a) Under the null hypothesis that there is no difference in the rate of evolution after duplication or speciation events, the distribution of phylogenetic independent contrasts would be drawn from the same distribution across both types of nodes. (b) Under the ortholog conjecture, contrasts across duplication nodes would tend to be larger than contrasts across speciation nodes."}

@@

For each internal node in each gene tree, we measured the phylogenetic independent contrast (PIC) of the expression statistic Tau. This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Figure 1b). These contrasts were then annotated by whether each is made across a speciation or duplication event. The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree. Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree.

@@

```{r show_contrast_result, echo=FALSE, cache=TRUE, message=F, warning=F, fig.cap="Density plot of the magnitude of phylogenetic independent contrast values following duplication and speciation events. As the contrasts for duplication events are not larger than those of speciation events, the results are not consistent with the predictions of the ortholog conjecture."}

@@

We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] datset onto `r length(gene_trees)` gene trees parsed from ENSMBL Compara [@Herrero:2016jj]. These are the same pre-computed trees that the orthology/ paralogy annotations KMRR used are based on. `r length(gene_trees_pruned)` passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values and had at least one speciation event. Of these, `r length(gene_trees_calibrated)` were successfully time calibrated. These calibrated trees were used to calculate phylogenetic independent contrasts for  `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes. One of these trees is shown in Supplementary Figure 7 to demonstrate the analysis.

It is essential to have a null hypothesis that makes a distinct prediction from the prediction of the hypothesis under consideration. A suitable null hypothesis in this case  is that there is no difference in the evolution of expression following speciation or duplication events [@Studer:2009bm]. Under this hypothesis, we would predict that contrasts across speciation nodes and duplication nodes are drawn from the same distribution (Figure 2a). Under the alternative hypothesis specified by the ortholog conjecture, that there is a higher rate of change following duplication events than speciation events, we would expect to see the distribution of duplication contrasts shifted to higher values relative to the speciation contrasts (Figure 2b).

We did not find increased evolutionary change in expression following duplication events (Figure 3). The Wilcoxon rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result`). Our phylogenetic comparative analysis, unlike the previously published pairwise comparative analysis [@KryuchkovaMostacci:2016iw], therefore find no evidence for the ortholog conjecture in this system.

@@

We next examined the possibility that ascertainment biases were differentially impacting the inference of expression evolution following duplication and speciation events. Such a bias might obscure support for the ortholog conjecture. We focused on two possible sources of bias, node depth and branch length. We found no evidence that either affected our results (Supplementary Information, Supplementary Figure 8). We also examined the sensitivity of the results to the calibration times applied to speciation events on the gene trees. This is important because it is expected that genes from separate species have a common ancestor older than the time at which the species diverged from each other [@Takahata:1989uh; @Degnan:2009hr]. There is also uncertainty associated with the timing of these speciation events. We added random noise to the calibration times in replicate analyses, and all still failed to reject the null hypothesis (Supplementary Information).

@@

```{r pairwise, echo=FALSE, cache=TRUE, message=F, warning=F, fig.cap="The pairwise analyses presented in Figure 2a of KMRR [@KryuchkovaMostacci:2016iw] can be reproduced with the subset of data we consider here. Each circle indicates the Tau Pearson correlation coefficient for a set of pairwise comparisons annotated with a specific node name of a given age and event type, *i.e.*, whether the divergence at that nodes was due to speciation (giving rise to orthologs) or duplication (giving rise to paralogs). Following KMRR Figure 2a, this figure shows only the comparisons that include a human sequence."}
	get_pairwise_summary = function ( nhx ) {

@@

	pairwise_summary %>% 
		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
		filter( ! is.na( D ) ) %>%
		group_by( Event, label ) %>% 
		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
		rename( clade = label ) %>% 
		left_join( calibration_times, clade=clade ) %>%
		ggplot() + 
			geom_point( aes( x=age, y=rho, col=Event ), size=3 ) +
			theme_classic() +
			xlab( "Node Age (million years)" ) +
			ylab( "Correlation Coefficient" )

@@

@@

Why is it that pairwise methods and phylogenetic methods lead to opposite conclusions? One reason is that multiple pairwise comparisons repeatedly sample the same evolutionary changes and in so doing violate statistical assumptions of independence, whereas phylogenetic comparative methods make multiple independent comparisons across non-overlapping branches of the tree.  The other reason for the different conclusions is that pairwise comparisons between the tips and phylogenetic comparative methods describe different things. Pairwise comparisons describe contemporary patterns, while phylogenetic methods infer historical processes [@OMeara:2012gq]. There need not be a different process of evolution following speciation and duplication for paralogs to be more different than orthologs. This could be due to the structure of the gene phylogenies. If paralogs tend to be more distantly related to each other than orthologs, then there would be more time for differences to accumulate even if the rate of change is the same between the two. This is, in fact, the case for these data. While the mean distance between orthologs is `r round( ortholog_distance_mean, 1 )` million years, the mean distance between paralogs is `r round( paralog_distance_mean, 1 )` million years. This is because the oldest speciation event is by definition the most recent common ancestor of the species included in the study, but many gene families underwent duplication before this time.

To test the hypothesis that distantly paralogs that arose from duplications that precede the oldest speciation event (Supplementary Figure 8a) impact the lower correlation of Tau between paralogs than between orthologs, we removed them. When we consider only the duplication events the same age or younger than the oldest speciation event (Supplementary Figure 8b), the paralog correlation coefficient increases from `r round( paralog_r, 3)` to  `r round( paralog_subset_r, 3)`. This is much closer to the ortholog correlation of `r round( ortholog_r, 3) `.

@@

```{r simulations_plot, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE, fig.cap="Analyses of simulated data reveal that pairwise methods produce similar results under both the null model and the ortholog conjecture, while the phylogenetic method distinguishes between the two. (a) Pairwise analysis of dataset simulated under the null model that there is no difference in the rate of expresison evolution following duplication and speciation. (b) Phylogenetic analysis of data simulated under the null model. (c) Pairwise analysis of dataset simulated under the ortholog conjecture, in which there is a higher rate of expresison evolution following duplication than speciation. This plot is very similar to the corresponding plot for the null model. (d) Phylogenetic analysis of data simulated under the ortholog conjecture. This plot is distinct from that for the null model in that phylogenetic independent contrasts for duplication events are significantly larger than after speciation events."}

@@

	g_pair_null = 
		pairwise_sim_null_summary %>% 
		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
		filter( ! is.na( D ) ) %>%
		group_by( Event, label ) %>% 
		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
		rename( clade = label ) %>% 
		left_join( calibration_times, clade=clade ) %>%
		ggplot() + 
			geom_point( aes( x=age, y=rho, col=Event ), size=3 ) +
			theme_classic() +
			xlab( "Node Age (million years)" ) +
			ylab( "Correlation Coefficient" ) +
			ylim( 0, 1 ) +
			ggtitle( "(a) Null simulation, pairwise" ) +
			theme(legend.position="none")

@@

	g_pair_oc = 
		pairwise_sim_oc_summary %>% 
		filter( species_a == "homo_sapiens" | species_b == "homo_sapiens" ) %>% 
		filter( ! is.na( D ) ) %>%
		group_by( Event, label ) %>% 
		summarise( cor.test(  tau_a, tau_b )$estimate, n=n() ) %>% 
		rename( rho=`cor.test(tau_a, tau_b)$estimate` ) %>%
		rename( clade = label ) %>% 
		left_join( calibration_times, clade=clade ) %>%
		ggplot() + 
			geom_point( aes( x=age, y=rho, col=Event ), size=3 ) +
			theme_classic() +
			xlab( "Node Age (million years)" ) +
			ylab( "Correlation Coefficient" ) +
			ylim( 0, 1 ) +
			ggtitle( "(c) OC simulation, pairwise" ) +
			theme(legend.position="none")

@@

The KMRR study did investigate the impact of node age on correlation, but in a different way. In their Figure 2, they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age. They found that across the investigated range of node ages, ortholog pairs have higher tau correlation than paralogs. We confirmed that we can replicate this result (Figure 4). There are several difficulty with interpreting this plot, though. First, it does not just reflect the evolutionary processes that generated the data, it is also impacted by the phylogenies along which these processes acted. The expected covariance of traits that evolve under neutral processes is in fact defined by the phylogeny [@Grafen:1989um]. Second, the correlation for each group is based on multiple non-independent pairwise comparisons.

To better understand this plot (Figure 4), we performed a couple simulations of Tau on the calibrated gene trees and then regenerated the figures. First, we simulated the evolution of tau under the null model that it evolves at the same rate following duplication and speciation events. Under the null model, the plot of correlation coefficient to node age is very similar as for the observed data (Figure 5a). As in the original study, there is higher correlation coefficient across orthologs (`r ortholog_r_sim_null`) than paralogs (`r paralog_r_sim_null`) when not considering node age. Phylogenetic analysis of the data simulated under the null hypothesis (Figure 5b) do not reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_null_contrast )`$), as expected.

We next simulated the evolution of Tau under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation. The pairwise results of this heterogeneous model are nearly indistinguishable from the null model (Figure 5c), and also have a higher correlation coefficient for orthologs (`r ortholog_r_sim_oc`) than paralogs (`r paralog_r_sim_oc`). The phylogenetic analysis of the ortholog conjecture simulation (Figure 5d) does reject the null hypothesis (Wilcoxon p = $`r wilcox_oc( nodes_sim_oc_contrast )`$).

@@

Their Figure 3 differs from their other analyses in that it presents independent changes in expression between triplets of genes. Each triplet has two paralogs that arose from a duplication event and one unduplicated homolog. They find that there is a tendency for the paralog with the lowest expression to have the highest Tau. From this they conclude that following duplication there is a common trajectory of evolutionary change, where one paralog evolves to have lower expression and also become more tissue specific. There is, however, a negative relationship between Tau and maximum expression across all genes (Supplementary Figure 9). A simpler explanation for this plot is that it reflects the global tendency for Tau to decrease with maximum expression. The null expectation for any two genes sampled at random is for one to have hhigher maximum expression and lower  This global pattern could be due to both biological factors and the technical details of how Tau is calculated.

@@

![Figure XXLevin | Distributions of pairwise similarity scores for each phase of development. Pairwise scores for the ctenophore are red. Wilcoxon test p-values for the significance of the differences between early-mid distributions and late-mid distributions are on the right. Model of variance, which is inversely related to similarity, is on the left. (a) The distributions as published. Low similarity (*i.e.*, high variance) in the mid phase of development was interpreted as support for an inverse hourglass model for the evolution of gene expression. The five least-similar mid phase scores were all from the ctenophore. Published KS p-values, based on duplicated data, are in parentheses. The inset ctenophore image is by S. Haddock from phylopic.org. (b) The distributions after the exclusion of the ctenophore. The early and mid phase distributions are not statistically distinct.](levin_etal/Figure_levin.png?raw=true)

@@

We examined the matrix of pairwise comparisons used as the base for the KS tests and Figure 4d in *Levin et al.* [@Levin:2016bp], and thus as evidence to support the "inverse hourglass model". We found several problems resulting from the use of multiple pairwise comparisons. The first problems are specific to this particular implementation of pairwise comparisons. We found that every data point was included twice because both reciprocal pairwise comparisons (which have the same values) were retained. For example, there is both a nematode to arthropod comparison and an arthropod to nematode comparison. As a consequence, there are 90 entries for the 45 pairwise comparisons, and by doubling the data the significance of the result appears stronger than it actually is. After removing the duplicate values, the p values are far less significant, 0.002 for the early-mid comparison and on the order of $10^-6$ for early-late. In addition, the test they used (KS test) is not appropriate for the hypothesis they seek to evaluate. The KS test does not just evaluate whether one distribution is greater than the other, it also tests whether the shape of the distributions are the same. In addition, the samples in this dataset are matched (*i.e.*, for each pairwise comparison there is a early, mid, and late expression value), which the KS test does not take into account. The Wilcoxon test is instead appropriate in this case. When applied to the de-duplicated data, the significance of this test is 0.02 for the early-mid comparison and on the order of $10^-7$ for early-late.
